Guilt, shame, and anti-war action in an authoritarian country at war

### 2024-06-12 manuscript.tex
Finished implementing revisions

\emph{Support for democratic values} was measured with a 6-item scale from Canetti-Nisim and Beit-Hallahmi (2007). Example item: ``Every citizen has the right to take his convictions to the street if necessary'' (\(\alpha\) = 0.69; M=\texttt{r}, SD=\texttt{r}).

@@

Since the constructs of political alienation, cynicism, and responsibility are closely related and none of these scales were used in a Russian sample before, we conducted a confirmatory factor analysis (CFA) to test whether the 3-factor structure could be confirmed. The 3-factor model showed unsatisfactory fit to the data (CFI = 0.803, RMSEA = 0.128, SRMR = 0.11). An exploratory factor analysis suggested that a 4-factor model, with the political cynicism scale split in two, fits the data better. CFA confirmed this result: a model with 4 factors showed a good fit to the data (CFI = 0.947, RMSEA = 0.068, SRMR = 0.068). This modified model includes the \emph{political responsibility} factor with 5 items (M=\texttt{r}, SD=\texttt{r}), \emph{political alienation} factor with 3 items (M=\texttt{r}, SD=\texttt{r}; the item ``I believe public officials don't care much what people like me think'' was removed since it loaded on both political alienation and cynicism subscales), and two facets of political cynicism: ``\emph{politicians are bad}'' (4 items, e.g., ``Politicians are only interested in getting and maintaining power''; M=\texttt{r}, SD=\texttt{r}) and ``\emph{politics is bad}'' (4 items, e.g., ``No man can hope to stay honest once he enters politics''; M=\texttt{r}, SD=\texttt{r}).

\textbf{\emph{Positive and negative emotions.}} We measured emotions that participants experience when thinking about the war using the short version of the Positive and Negative Affect Scale {[}PANAS; Watson et al. (1988){]}. Twelve emotions (fear, joy, sadness, hope, anger, enthusiasm, disgust, pride, contempt, depression, guilt, and shame) were rated on a 5-point scale from 1 (\emph{do not experience at all}) to 5 (\emph{extremely}). If participants answered 2 or higher to the guilt and shame questions, they were shown questionnaires for group-based guilt (n=) and/or shame (n=), respectively. We took this precaution participants in earlier studies conducted in Russia complained about the double-barreled nature of items measuring group-based guilt and shame, where they felt that the questions imply that participants experience some level of guilt or shame (see Grigoryan et al. (2018); Grigoryan and Efremova (2017), for details). We tested the two-factor structure of the PANAS scale (positive and negative affect, excluding guilt and shame\footnote{We excluded guilt and shame since we planned to estimate whether group-based guilt and shame are predictive of behavior over and above other emotions. Including PANAS measures of guilt and shame into the index of negative emotions would have confounded the two measures and prevented us from addressing this question (RQ1).}) using CFA and found the fit to be unsatisfactory (CFI = 0.862, RMSEA = 0.146, SRMR = 0.08). A 3-factor structure with \emph{positive emotions} (joy, hope, enthusiasm; M=, SD=), \emph{negative dominant emotions} (anger, disgust, contempt; M=, SD=) and \emph{negative submissive emotions} (fear, sadness, depression; M=, SD=) showed a satisfactory fit to the data (CFI = 0.916, RMSEA = 0.117, SRMR = 0.066; see Mehrabian (1980) for the distinction between dominant and submissive emotions).

\textbf{\emph{Group-based guilt and shame.}} We used the measures of group-based guilt, image shame, and moral shame developed by Rees et al. (2013), Russian adaptation by Grigoryan and Efremova (2017). All items were answered on a 5-point scale from 1 (\emph{completely disagree}) to 5 (\emph{completely agree}). Guilt was measured with 3 items (e.g., ``I feel guilty for the manner in which Ukrainian people have been treated by Russians'', \(\alpha\) = 0.89), image shame with 5 items (``I feel ashamed when I realize that other countries might think of Russia negatively because of our involvement in Ukraine'', \(\alpha\) = 0.94), and moral shame with 4 items (``I feel ashamed because Russia's actions with regard to Ukraine have been immoral'', \(\alpha\) = 0.93). All missing values were replaced with zeros, since participants who didn't see the scale were the ones who reported no feelings of guilt or shame in the PANAS. With this imputation, we were able to use the whole dataset without excluding those who reported experiencing no guilt or shame. Mean values after this imputation were M=1.23 (SD=1.69) for guilt, M=1.13 (SD=1.70) for image shame, and M=1.09 (SD=1.67) for moral shame.

@@

\textbf{\emph{Anti-war actions}}. We measured \emph{past behavior} by asking how many of the eight political actions listed the participant had done to oppose the war in Ukraine (e.g., ``signing a petition'', ``participating in a demonstration'', etc.). The responses ranged from 0 to 8, with 82\% of participants reporting zero actions, M=, SD=. \emph{Behavioral intention} to participate in any political action against the war was measured with 3 items (``I want to/intend to/plan to take part in political action against the war in Ukraine''), on a 5-point scale from 1 (\emph{completely disagree}) to 5 (\emph{completely agree}) (\(\alpha\) = 0.95; 73\% of participants had a mean score of 1; M=, SD=). Finally, we asked about \emph{the probability that the person will take any political action}, from 1 (\emph{highly improbable}), 10 (\emph{(almost) certainly}) (69\% answered `1'; M=, SD=).

@@

{\centering \includegraphics{corplot}

@@

\begin{table}[H]
\centering
\caption{\label{tab:Table1}Predicting group-based guilt and shame from political beliefs}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lrrrrrrrrrrrr}
\toprule
\multicolumn{1}{c}{} & \multicolumn{4}{c}{Moral Shame} & \multicolumn{4}{c}{Image Shame} & \multicolumn{4}{c}{Guilt} \\
\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9} \cmidrule(l{3pt}r{3pt}){10-13}
\em{ } & \em{$\beta$} & \em{b} & \em{SE} & \em{p} & \em{$\beta$} & \em{b} & \em{SE} & \em{p} & \em{$\beta$} & \em{b} & \em{SE} & \em{p}\\
\midrule
Intercept & 0 & -2.89 & 0.32 & <.001 & 0 & -2.85 & 0.34 & <.001 & 0 & -2.93 & 0.34 & <.001\\
Democratic Values & 0.41 & 0.64 & 0.05 & <.001 & 0.4 & 0.64 & 0.05 & <.001 & 0.41 & 0.65 & 0.05 & <.001\\
Political Cynicism (politicians) & 0.18 & 0.2 & 0.05 & <.001 & 0.16 & 0.18 & 0.05 & <.001 & 0.09 & 0.1 & 0.05 & .039\\
Political Cynicism (politics) & -0.05 & -0.05 & 0.04 & .189 & -0.03 & -0.03 & 0.04 & .458 & -0.03 & -0.03 & 0.04 & .455\\
Political Alienation & -0.05 & -0.06 & 0.04 & .122 & -0.03 & -0.04 & 0.04 & .380 & 0.01 & 0.02 & 0.04 & .663\\
\addlinespace
Political Responsibility & 0.04 & 0.05 & 0.04 & .190 & 0.03 & 0.04 & 0.04 & .347 & 0.07 & 0.1 & 0.04 & .022\\
\midrule\\
R$^{2}$ &  & .260 &  &  &  & .234 &  &  &  & .221 &  & \\
\bottomrule
\end{tabular}
\end{table}

@@

In a series of linear regressions, we tested whether guilt, image shame, and moral shame are predictive of anti-war actions in the past and in the future. Only moral shame was consistently related to anti-war action and the results were consistent in linear (Table \ref{tab:Table2}) and zero-inflated negative binomial models (Table \ref{tab:TableS3}, SM). Group-based guilt and shame explained about 18\% of variance in past behavior and about 40\% of variance in future behavior. Due to strong correlations between moral shame and both image shame (\emph{r} = .94***) and guilt (\emph{r} = .85***), these models showed a multicollinearity problem with VIF of \(\approx 9\) for image shame and of \(\approx 11\) for moral shame across models. We therefore additionally tested how much variance each of the predictors explain in behavioral outcomes when each of them is the only predictor in the model. Across all three outcomes, moral shame explained most variance (18\% for past behavior, 42\% for intention, and 37\% for action probability), followed by image shame (15\%, 39\%, 33\%), and then guilt (13\%, 31\%, 28\%).

\begin{table}[H]
\centering
\caption{\label{tab:Table2}Predicting anti-war action from group-based guilt and shame}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lrrrrrrrrrrrr}
\toprule
\multicolumn{1}{c}{} & \multicolumn{4}{c}{Past Behaviour} & \multicolumn{4}{c}{Behavioural Intention} & \multicolumn{4}{c}{Action Probability} \\
\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9} \cmidrule(l{3pt}r{3pt}){10-13}
\em{ } & \em{$\beta$} & \em{b} & \em{SE} & \em{p} & \em{$\beta$} & \em{b} & \em{SE} & \em{p} & \em{$\beta$} & \em{b} & \em{SE} & \em{p}\\
\midrule
Intercept & 0 & 0.1 & 0.04 & .009 & 0 & 1.09 & 0.03 & <.001 & 0 & 1.21 & 0.07 & <.001\\
Guilt & 0.02 & 0.01 & 0.04 & .756 & 0 & 0 & 0.03 & .942 & 0.05 & 0.06 & 0.06 & .326\\
Image Shame & -0.07 & -0.05 & 0.05 & .393 & 0.1 & 0.06 & 0.04 & .154 & -0.02 & -0.02 & 0.1 & .826\\
Moral Shame & 0.48 & 0.31 & 0.06 & <.001 & 0.56 & 0.33 & 0.05 & <.001 & 0.59 & 0.77 & 0.11 & <.001\\
\midrule\\
R$^{2}$ &  & .181 &  &  &  & .425 &  &  &  & .376 &  & \\
\bottomrule
\end{tabular}
\end{table}

@@

Similar to the previous analysis, moral shame showed a high VIF of 11.9. We tested the model excluding moral shame as a predictor. There was no multicollinearity problem detected after exclusion of moral shame, all VIFs \textless{} 4. In this revised model, negative dominant emotions were the strongest predictor of future behavior (\(\beta\) = .28, \emph{p} \textless{} .001), followed by image shame (\(\beta\) = .24, \emph{p} \textless{} .001). Importantly, moral shame alone explained almost as much variance in future behavior (43\%) as all the other predictors combined (46\%). None of the other predictors individually explained as much variance as moral shame.

\begin{table}[H]
\centering
\caption{\label{tab:Table3}Predicting future behavior from emotions and attitudes
}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lrrrrr}
\toprule
\em{ } & \em{$\beta$} & \em{b} & \em{SE} & \em{t-value} & \em{p}\\
\midrule
Intercept & 0 & -0.1 & 0.12 & -0.82 & .410\\
Emotions: Positive & -0.032 & -0.03 & 0.03 & -0.96 & .337\\
Emotions: Negative Dominant & 0.237 & 0.19 & 0.03 & 6.4 & <.001\\
Emotions: Negative Submissive & -0.076 & -0.07 & 0.03 & -2.43 & .015\\
Attitudes towards the war & -0.152 & -0.12 & 0.03 & -3.77 & <.001\\
\addlinespace
Group-based moral shame & 0.36 & 0.22 & 0.05 & 4.63 & <.001\\
Group-based image shame & 0.002 & 0 & 0.04 & 0.03 & .977\\
Group-based guilt & 0.033 & 0.02 & 0.03 & 0.76 & .450\\
\bottomrule
\end{tabular}
\end{table}

@@

Finally, since the measures of group-based guilt and shame confound emotions and appraisals, we additionally tested the predictive power of guilt and shame for future anti-war actions using the PANAS measure that captures only emotions, but not appraisals. We additionally included anger in this model, since anger has been shown to motivate reconciliatory actions on par or even more strongly than guilt and shame (Hakim et al., 2021; Iyer et al., 2007). The model showed no multicollinearity issues (all VIFs \textless{} 4). Shame was the strongest predictor of future anti-war actions (\(\beta\) = .46, \emph{p} \textless{} .001), followed by anger (\(\beta\) = .20, \emph{p} \textless{} .001). Guilt did not predict behavior once shame and anger were controlled (\(\beta\) = .04, \emph{p} = .440). Similarly, when anger was included in the model alongside the three group-based emotions, moral shame was once again the strongest predictor (\(\beta\) = .53, \emph{p} \textless{} .001), followed by anger (\(\beta\) = .14, \emph{p} \textless{} .001). Image shame (\(\beta\) = .03, \emph{p} = .680) and guilt (\(\beta\) = .02, \emph{p} = .620) had no effect.

@@

We were able to test this prediction with our current dataset, as it included a measure of political alienation -- the belief that ordinary people have no influence over the actions of their governments. Both the effects of guilt and moral shame on future anti-war action were moderated by political alienation: the less participants believed that people have control over the actions of their governments, the weaker was the effect of both guilt (\(\beta\) = -.169, SE = .03, \emph{p} \textless{} .001) and moral shame (\(\beta\) = -.165, SE = .02, \emph{p} \textless{} .001) on future anti-war action. However, the effect sizes of these interactions were nearly identical, suggesting that this cannot explain the lack of the main effect for guilt. We could speculate that the effects of guilt and shame have more to do with appraisals than with the emotion itself: moral outrage could be the main driving force behind the effects of guilt, shame, and anger on reconciliatory intergroup behavior found in various studies, and once this appraisal is accounted for, there is no added predictive value in feeling guilt.

@@

\begin{table}[H]
\centering
\caption{\label{tab:TableS3}Predicting anti-war action from group-based shame and guilt}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lrrrrrrrrr}
\toprule
\multicolumn{1}{c}{\textbf{}} & \multicolumn{3}{c}{\textbf{Past behavior}} & \multicolumn{3}{c}{\textbf{Behavioral inention}} & \multicolumn{3}{c}{\textbf{Action probability}} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-10}
\em{ } & \em{b} & \em{SE} & \em{p} & \em{b} & \em{SE} & \em{p} & \em{b} & \em{SE} & \em{p}\\
\midrule
\em{\textbf{Count model}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}}\\
\midrule\\
Intercept & 0.65 & 0.16 & <.001 & -0.22 & 0.15 & .142 & 0.39 & 0.1 & <.001\\
Guilt & 0 & 0.07 & .953 & -0.03 & 0.05 & .530 & 0.02 & 0.04 & .606\\
Image shame & -0.09 & 0.09 & .313 & 0 & 0.08 & .988 & 0.02 & 0.06 & .671\\
Moral shame & 0.08 & 0.1 & .451 & 0.25 & 0.08 & .003 & 0.22 & 0.07 & <.001\\
\midrule\\
\addlinespace
\em{\textbf{Zero-inflation model}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}}\\
\midrule\\
Intercept & 2.72 & 0.19 & <.001 & 1.7 & 0.2 & <.001 & 1.4 & 0.14 & <.001\\
Guilt & -0.05 & 0.13 & .696 & -0.18 & 0.13 & .156 & -0.2 & 0.09 & .033\\
Image shame & -0.18 & 0.17 & .295 & -0.47 & 0.26 & .076 & -0.13 & 0.14 & .384\\
Moral shame & -0.65 & 0.19 & <.001 & -0.26 & 0.27 & .337 & -0.35 & 0.16 & .033\\
\bottomrule
\end{tabular}
\end{table}

>>>>

\emph{Support for democratic values} was measured with a 6-item scale from Canetti-Nisim and Beit-Hallahmi (2007). Example item: ``Every citizen has the right to take his convictions to the street if necessary'' (\(\alpha\) = 0.69; M=4.99, SD=1.07).

@@

Since the constructs of political alienation, cynicism, and responsibility are closely related and none of these scales were used in a Russian sample before, we conducted a confirmatory factor analysis (CFA) to test whether the 3-factor structure could be confirmed. The 3-factor model showed unsatisfactory fit to the data (CFI = 0.803, RMSEA = 0.128, SRMR = 0.11). An exploratory factor analysis suggested that a 4-factor model, with the political cynicism scale split in two, fits the data better. CFA confirmed this result: a model with 4 factors showed a good fit to the data (CFI = 0.947, RMSEA = 0.068, SRMR = 0.068). This modified model includes the \emph{political responsibility} factor with 5 items (M=4.32, SD=1.15), \emph{political alienation} factor with 3 items (M=4.62, SD=1.38; the item ``I believe public officials don't care much what people like me think'' was removed since it loaded on both political alienation and cynicism subscales), and two facets of political cynicism: ``\emph{politicians are bad}'' (4 items, e.g., ``Politicians are only interested in getting and maintaining power''; M=5.2, SD=1.5) and ``\emph{politics is bad}'' (4 items, e.g., ``No man can hope to stay honest once he enters politics''; M=4.1, SD=1.52).

\textbf{\emph{Positive and negative emotions.}} We measured emotions that participants experience when thinking about the war using the short version of the Positive and Negative Affect Scale {[}PANAS; Watson et al. (1988){]}. Twelve emotions (fear, joy, sadness, hope, anger, enthusiasm, disgust, pride, contempt, depression, guilt, and shame) were rated on a 5-point scale from 1 (\emph{do not experience at all}) to 5 (\emph{extremely}). If participants answered 2 or higher to the guilt and shame questions, they were shown questionnaires for group-based guilt (\emph{n} = 400) and/or shame (\emph{n} = 349), respectively. We took this precaution participants in earlier studies conducted in Russia complained about the double-barreled nature of items measuring group-based guilt and shame, where they felt that the questions imply that participants experience some level of guilt or shame (see Grigoryan et al. (2018); Grigoryan and Efremova (2017), for details). We tested the two-factor structure of the PANAS scale (positive and negative affect, excluding guilt and shame\footnote{We excluded guilt and shame since we planned to estimate whether group-based guilt and shame are predictive of behavior over and above other emotions. Including PANAS measures of guilt and shame into the index of negative emotions would have confounded the two measures and prevented us from addressing this question (RQ1).}) using CFA and found the fit to be unsatisfactory (CFI = 0.862, RMSEA = 0.146, SRMR = 0.08). A 3-factor structure with \emph{positive emotions} (joy, hope, enthusiasm; M=2.19, SD=1.04), \emph{negative dominant emotions} (anger, disgust, contempt; M=2.34, SD=1.25) and \emph{negative submissive emotions} (fear, sadness, depression; M=2.88, SD=1.11) showed a satisfactory fit to the data (CFI = 0.916, RMSEA = 0.117, SRMR = 0.066; see Mehrabian (1980) for the distinction between dominant and submissive emotions).

\textbf{\emph{Group-based guilt and shame.}} We used the measures of group-based guilt, image shame, and moral shame developed by Rees et al. (2013), Russian adaptation by Grigoryan and Efremova (2017). All items were answered on a 5-point scale from 1 (\emph{completely disagree}) to 5 (\emph{completely agree}). Guilt was measured with 3 items (e.g., ``I feel guilty for the manner in which Ukrainian people have been treated by Russians'', \(\alpha\) = 0.89), image shame with 5 items (``I feel ashamed when I realize that other countries might think of Russia negatively because of our involvement in Ukraine'', \(\alpha\) = 0.94), and moral shame with 4 items (``I feel ashamed because Russia's actions with regard to Ukraine have been immoral'', \(\alpha\) = 0.93). All missing values were replaced with zeros, since participants who didn't see the scale were the ones who reported no feelings of guilt or shame in the PANAS. With this imputation, we were able to use the whole dataset without excluding those who reported experiencing no guilt or shame. Mean values after this imputation were M=1.23 (SD=1.69) for guilt, M=1.13 (SD=1.7) for image shame, and M=1.09 (SD=1.67) for moral shame.

@@

\textbf{\emph{Anti-war actions}}. We measured \emph{past behavior} by asking how many of the eight political actions listed the participant had done to oppose the war in Ukraine (e.g., ``signing a petition'', ``participating in a demonstration'', etc.). The responses ranged from 0 to 8, with 82\% of participants reporting zero actions, M=0.4, SD=1.07. \emph{Behavioral intention} to participate in any political action against the war was measured with 3 items (``I want to/intend to/plan to take part in political action against the war in Ukraine''), on a 5-point scale from 1 (\emph{completely disagree}) to 5 (\emph{completely agree}) (\(\alpha\) = 0.95; 73\% of participants had a mean score of 1; M=1.53, SD=1). Finally, we asked about \emph{the probability that the person will take any political action}, from 1 (\emph{highly improbable}), 10 (\emph{(almost) certainly}) (69\% answered `1'; M=2.09, SD=2.17).

@@

{\centering \includegraphics{outputs/corplot}

@@

\begin{table}[H]
\centering
\caption{\label{tab:Table1}Predicting group-based guilt and shame from political beliefs}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lrrrrrr}
\toprule
  & $\beta$ & lower 95\%CI & upper 95\%CI & $b$ & SE & $p$\\
\midrule
\textbf{Moral shame} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
\midrule
Intercept &  &  &  & -2.89 & 0.32 & <.001\\
Democratic Values & 0.41 & 0.35 & 0.48 & 0.64 & 0.05 & <.001\\
Political Cynicism (politicians) & 0.18 & 0.1 & 0.26 & 0.2 & 0.05 & <.001\\
Political Cynicism (politics) & -0.05 & -0.12 & 0.02 & -0.05 & 0.04 & .189\\
Political Alienation & -0.05 & -0.11 & 0.01 & -0.06 & 0.04 & .122\\
Political Responsibility & 0.04 & -0.02 & 0.09 & 0.05 & 0.04 & .190\\
\midrule\\
\textbf{Image shame} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
\midrule
Intercept &  &  &  & -2.85 & 0.34 & <.001\\
Democratic Values & 0.4 & 0.33 & 0.46 & 0.64 & 0.05 & <.001\\
Political Cynicism (politicians) & 0.16 & 0.07 & 0.24 & 0.18 & 0.05 & <.001\\
Political Cynicism (politics) & -0.03 & -0.1 & 0.05 & -0.03 & 0.04 & .458\\
Political Alienation & -0.03 & -0.09 & 0.04 & -0.04 & 0.04 & .380\\
Political Responsibility & 0.03 & -0.03 & 0.08 & 0.04 & 0.04 & .347\\
\midrule\\
\textbf{Guilt} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
\midrule
Intercept &  &  &  & -2.93 & 0.34 & <.001\\
Democratic Values & 0.41 & 0.35 & 0.48 & 0.65 & 0.05 & <.001\\
Political Cynicism (politicians) & 0.09 & 0 & 0.18 & 0.1 & 0.05 & .039\\
Political Cynicism (politics) & -0.03 & -0.1 & 0.05 & -0.03 & 0.04 & .455\\
Political Alienation & 0.01 & -0.05 & 0.08 & 0.02 & 0.04 & .663\\
Political Responsibility & 0.07 & 0.01 & 0.13 & 0.1 & 0.04 & .022\\
\bottomrule
\multicolumn{7}{l}{\textsuperscript{} Note: R² moral shame = .260, R² image shame = .234, R² guilt = .221}\\
\end{tabular}
\end{table}

@@

In a series of linear regressions, we tested whether guilt, image shame, and moral shame are predictive of anti-war actions in the past and in the future. Only moral shame was consistently related to anti-war action and the results were consistent in linear (Table \ref{tab:Table2}) and zero-inflated negative binomial models (Table \ref{tab:TableS3}, SM). Group-based guilt and shame explained about 18\% of variance in past behavior and about 40\% of variance in future behavior. Due to strong correlations between moral shame and both image shame (\emph{r} = 0.94***) and guilt (\emph{r} = 0.85***), these models showed a multicollinearity problem with VIF of \(\approx 9\) for image shame and of \(\approx 11\) for moral shame across models. We therefore additionally tested how much variance each of the predictors explain in behavioral outcomes when each of them is the only predictor in the model. Across all three outcomes, moral shame explained most variance (18\% for past behavior, 42\% for intention, and 38\% for action probability), followed by image shame (15\%, 39\%, 33\%), and then guilt (13\%, 31\%, 29\%).

\begin{table}[H]
\centering
\caption{\label{tab:Table2}Predicting anti-war action from group-based guilt and shame}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lrrrrrr}
\toprule
  & $\beta$ & lower 95\%CI & upper 95\%CI & $b$ & SE & $p$\\
\midrule
\textbf{Past behavior} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
\midrule
Intercept &  &  &  & 0.1 & 0.04 & .009\\
Guilt & 0.02 & -0.09 & 0.13 & 0.01 & 0.04 & .756\\
Image Shame & -0.07 & -0.24 & 0.1 & -0.05 & 0.05 & .393\\
Moral Shame & 0.48 & 0.29 & 0.67 & 0.31 & 0.06 & <.001\\
\midrule\\
\textbf{Behavioral intention} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
\midrule
Intercept &  &  &  & 1.09 & 0.03 & <.001\\
Guilt & 0 & -0.1 & 0.09 & 0 & 0.03 & .942\\
Image Shame & 0.1 & -0.04 & 0.25 & 0.06 & 0.04 & .154\\
Moral Shame & 0.56 & 0.4 & 0.71 & 0.33 & 0.05 & <.001\\
\midrule\\
\textbf{Action probability} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
\midrule
Intercept &  &  &  & 1.21 & 0.07 & <.001\\
Guilt & 0.05 & -0.05 & 0.14 & 0.06 & 0.06 & .326\\
Image Shame & -0.02 & -0.16 & 0.13 & -0.02 & 0.1 & .826\\
Moral Shame & 0.59 & 0.42 & 0.75 & 0.77 & 0.11 & <.001\\
\bottomrule
\multicolumn{7}{l}{\textsuperscript{} Note: R² past behavior = .181, R² behavioral intention = .425, R² action probability = .376}\\
\end{tabular}
\end{table}

@@

Similar to the previous analysis, moral shame showed a high VIF of 11.9. We tested the model excluding moral shame as a predictor. There was no multicollinearity problem detected after exclusion of moral shame, all VIFs \textless{} 4. In this revised model, negative dominant emotions were the strongest predictor of future behavior (\(\beta\) = 0.28, \emph{p} \textless.001), followed by image shame (\(\beta\) = 0.24, \emph{p} \textless.001). Importantly, moral shame alone explained almost as much variance in future behavior (43\%) as all the other predictors combined (47\%). None of the other predictors individually explained as much variance as moral shame.

\begin{table}[H]
\centering
\caption{\label{tab:Table3}Predicting future behavior from emotions and attitudes
}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{llrrrrrr}
\toprule
\em{ } & \em{$\beta$} & \em{lower 95\%CI} & \em{upper 95\%CI} & \em{$b$} & \em{$SE$} & \em{$t-value$} & \em{$p$}\\
\midrule
Intercept &  &  &  & -0.1 & 0.12 & -0.82 & .410\\
Emotions: Positive & -0.032 & -0.1 & 0.03 & -0.03 & 0.03 & -0.96 & .337\\
Emotions: Negative Dominant & 0.237 & 0.16 & 0.31 & 0.19 & 0.03 & 6.4 & <.001\\
Emotions: Negative Submissive & -0.076 & -0.14 & -0.01 & -0.07 & 0.03 & -2.43 & .015\\
Attitudes towards the war & -0.152 & -0.23 & -0.07 & -0.12 & 0.03 & -3.77 & <.001\\
Group-based moral shame & 0.36 & 0.21 & 0.51 & 0.22 & 0.05 & 4.63 & <.001\\
Group-based image shame & 0.002 & -0.13 & 0.13 & 0 & 0.04 & 0.03 & .977\\
Group-based guilt & 0.033 & -0.05 & 0.12 & 0.02 & 0.03 & 0.76 & .450\\
\bottomrule
\end{tabular}
\end{table}

@@

Finally, since the measures of group-based guilt and shame confound emotions and appraisals, we additionally tested the predictive power of guilt and shame for future anti-war actions using the PANAS measure that captures only emotions, but not appraisals. We additionally included anger in this model, since anger has been shown to motivate reconciliatory actions on par or even more strongly than guilt and shame (Hakim et al., 2021; Iyer et al., 2007). The model showed no multicollinearity issues (all VIFs \textless{} 4). Shame was the strongest predictor of future anti-war actions (\(\beta\) = 0.46, \emph{p} \textless.001), followed by anger (\(\beta\) = 0.2, \emph{p} \textless.001). Guilt did not predict behavior once shame and anger were controlled (\(\beta\) = 0.04, \emph{p} = .436). Similarly, when anger was included in the model alongside the three group-based emotions, moral shame was once again the strongest predictor (\(\beta\) = 0.53, \emph{p} \textless.001), followed by anger (\(\beta\) = 0.15, \emph{p} \textless.001). Image shame (\(\beta\) = 0.03, \emph{p} = .679) and guilt (\(\beta\) = 0.02, \emph{p} = .621) had no effect.

@@

We were able to test this prediction with our current dataset, as it included a measure of political alienation -- the belief that ordinary people have no influence over the actions of their governments. Both the effects of guilt and moral shame on future anti-war action were moderated by political alienation: the less participants believed that people have control over the actions of their governments, the weaker was the effect of both guilt (\(\beta\) = -0.169, SE = 0.03, \emph{p} \textless.001) and moral shame (\(\beta\) = -0.165, SE = 0.02, \emph{p} \textless.001) on future anti-war action. However, the effect sizes of these interactions were nearly identical, suggesting that this cannot explain the lack of the main effect for guilt. We could speculate that the effects of guilt and shame have more to do with appraisals than with the emotion itself: moral outrage could be the main driving force behind the effects of guilt, shame, and anger on reconciliatory intergroup behavior found in various studies, and once this appraisal is accounted for, there is no added predictive value in feeling guilt.

@@

\begin{table}[H]
\centering
\caption{\label{tab:TableS3}Predicting anti-war action from group-based shame and guilt}
\centering
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{lrrrrrrrrr}
\toprule
\multicolumn{1}{c}{\textbf{}} & \multicolumn{3}{c}{\textbf{Past behavior}} & \multicolumn{3}{c}{\textbf{Behavioral intention}} & \multicolumn{3}{c}{\textbf{Action probability}} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-10}
\em{ } & \em{b} & \em{SE} & \em{p} & \em{b} & \em{SE} & \em{p} & \em{b} & \em{SE} & \em{p}\\
\midrule
\em{\textbf{Count model}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}}\\
\midrule\\
Intercept & 0.65 & 0.16 & <.001 & -0.22 & 0.15 & .142 & 0.39 & 0.1 & <.001\\
Guilt & 0 & 0.07 & .953 & -0.03 & 0.05 & .530 & 0.02 & 0.04 & .606\\
Image shame & -0.09 & 0.09 & .313 & 0 & 0.08 & .988 & 0.02 & 0.06 & .671\\
Moral shame & 0.08 & 0.1 & .451 & 0.25 & 0.08 & .003 & 0.22 & 0.07 & <.001\\
\midrule\\
\addlinespace
\em{\textbf{Zero-inflation model}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}} & \em{\textbf{}}\\
\midrule\\
Intercept & 2.72 & 0.19 & <.001 & 1.7 & 0.2 & <.001 & 1.4 & 0.14 & <.001\\
Guilt & -0.05 & 0.13 & .696 & -0.18 & 0.13 & .156 & -0.2 & 0.09 & .033\\
Image shame & -0.18 & 0.17 & .295 & -0.47 & 0.26 & .076 & -0.13 & 0.14 & .384\\
Moral shame & -0.65 & 0.19 & <.001 & -0.26 & 0.27 & .337 & -0.35 & 0.16 & .033\\
\bottomrule
\end{tabular}
\end{table}
